Inference of proto-neutron star properties from gravitational-wave data in core-collapse supernovae.

### 2020-11-07 introduction.tex
Further work on the paper.

The life of sufficiently massive stars, i.e.~those born with masses between $\sim 8$~M$_\odot$ and $\sim 120$~M$_\odot$, ends with the collapse of {the} iron core under {its} own gravity, leading {to} the formation of a neutron star {(NS)} or a black hole (BH), {and} followed (typically but not necessarily in the BH case) by {a supernova} explosion. Nearby core-collapse supernova (CCSN) explosions are expected to be sources of gravitational waves (GWs) and they could be 
%are one of the main candidates for 
the next great discovery of current ground-based observatories. However, these are relative rare events. A neutrino-driven explosion \citep{Bethe:1990} is the most likely outcome in the case of slowly rotating cores, which are present in the bulk of CCSN progenitors. The emitted GWs could be detected with the advanced ground-based GW detector network (Advanced LIGO~\citep{TheLIGOScientific:2014jea}, Advanced Virgo~\citep{TheVirgo:2014hva} and
KAGRA~\citep{Aso:2013eba}) within $5$~kpc \citep{Gossan:2016,TargetedSNSearchO12}. Such a galactic event has a rate of about $2-3$ per century \citep{Adams:2013,Rozwadowska:2021}.
For the case of rapidly rotating progenitor cores the result is likely a magneto-rotational explosion, yielding  a more powerful GW signal that could be detected within $50$~kpc and, for some extreme models, up to $5-30$ Mpc \citep{Gossan:2016,TargetedSNSearchO12}. However, only about $1\%$ of the electromagnetically observed events show signatures of fast rotation (broad-lined type Ic SNe \citep{Li:2011b} or events associated with long GRBs 
\citep{Chapman:2007}) making this possibility a subdominant channel of detection with an event rate of $\sim 10^{-4} \rm{yr}^{-1}$\mab{[add ref?]}. For the results discussed in this work we only consider neutrino-driven CCSN. \tf{This last sentence must be mentioned somewhere (or maybe not) but perhaps here is not the best place.} Despite the low rates, CCSN are of great scientific interest because they produce complex GW signals which could provide significant clues about the physical processes at work after the gravitational collapse of stellar cores.

>>>>

The life of sufficiently massive stars, i.e.~those born with masses between $\sim 8$~M$_\odot$ and $\sim 120$~M$_\odot$, ends with the collapse of {the} iron core under {its} own gravity, leading {to} the formation of a neutron star {(NS)} or a black hole (BH), {and} followed (typically but not necessarily in the BH case) by {a supernova} explosion. Nearby core-collapse supernova (CCSN) explosions are expected to be sources of gravitational waves (GWs) and they could be 
%are one of the main candidates for 
the next great discovery of current ground-based observatories. However, these are relative rare events. A neutrino-driven explosion \citep{Bethe:1990} is the most likely outcome in the case of slowly rotating cores, which are present in the bulk of CCSN progenitors. The emitted GWs could be detected with the advanced ground-based GW detector network (Advanced LIGO (aLIGO)~\citep{TheLIGOScientific:2014jea}, Advanced Virgo (AdV)~\citep{TheVirgo:2014hva} and
KAGRA~\citep{Aso:2013eba}) within $5$~kpc \citep{Gossan:2016,TargetedSNSearchO12}. Such a galactic event has a rate of about $2-3$ per century \citep{Adams:2013,Rozwadowska:2021}.
For the case of rapidly rotating progenitor cores the result is likely a magneto-rotational explosion, yielding  a more powerful GW signal that could be detected within $50$~kpc and, for some extreme models, up to $5-30$ Mpc \citep{Gossan:2016,TargetedSNSearchO12}. However, only about $1\%$ of the electromagnetically observed events show signatures of fast rotation (broad-lined type Ic SNe \citep{Li:2011b} or events associated with long GRBs 
\citep{Chapman:2007}) making this possibility a subdominant channel of detection with an event rate of $\sim 10^{-4} \rm{yr}^{-1}$\mab{[add ref?]}. For the results discussed in this work we only consider neutrino-driven CCSN. \tf{This last sentence must be mentioned somewhere (or maybe not) but perhaps here is not the best place.} Despite the low rates, CCSN are of great scientific interest because they produce complex GW signals which could provide significant clues about the physical processes at work after the gravitational collapse of stellar cores.

### 2020-11-07 results.tex
Further work on the paper.

To estimate how accurately we can infer the time evolution of $r=M_{\rm PNS}/R_{\rm PNS}^2$ in a single
detector GW data, we have added the GW signal {\texttt s20S} to 
100 Gaussian noise realisations whose power spectral density follows the Advanced LIGO (aLIGO)
spectrum~\cite{aLIGOsens:2018} shown on Figure~\ref{fig:spectrum}.

\begin{figure}
 \centering
 \includegraphics[width=0.5\textwidth]{plots/spectrum}
 \caption{Amplitude spectral density of the GW detectors Advanced LIGO (aLIGO) and Advanced Virgo (ADV) at design sensitivity and of the proposed third generation detectors Cosmic Explorer and Einstein Telescope. Eintein Telescope sensitivity curve ET-B is obtained pushing second generation detector technology at its limit. ET-C and ET-D sensitivity curves correspond to a detector configuration where a low-power cryogenic low-frequency interferometer and a high-power room temperature high-frequency interferometer are sharing the same infrastructure~\cite{Hild_2011}. Cosmic Explorer design sensitivity will be achieved in two stages. Stage 1 (CE1) is expected to use the technology developed for the “A+” upgrade to Advanced LIGO but scaled up to a 40 km detector while stage 2 (CE2) will implement state-of-the-art technology to decrease quantum and thermal noises~\cite{reitze2019cosmic}. } \label{fig:spectrum}
\end{figure}

We have covered a large range of distances for which a detection in second generation of GW
detectors is feasible. The source is optimally oriented with
respect to the GW single detector. We are assuming a GW signal from a core collapse
phenomena has been identified in the data and that the beginning of the GW signal is known within $O(10~ms)$.
The data (signal embedded in noise) are whitened using the function {\it prewhiten} of the R-package TSA.
An auto-regressive model with maximal \textcolor{red}{100} coefficients has been used.

For each of the noise realizations, we reconstruct the ratio time series {$r_i$}
of length $N$ starting from the left side of the spectrogram and constraining the beginning of the
track to be smaller than \unit[200]{Hz}. The reconstructed ratio is then compared to the ``true'' ratio
{$r_i^0$} derived from the PNS mass and radius computed from the {\texttt s20S} simulation.

Figure \ref{fig:s20results} shows the distribution of the fraction of the ratio {$r_i^0$} values
that fall within the 95\% confidence interval of {$r_i$}. This quantity, $coverage$, is taking maximal values
when the source is located within few kpc and then decreases with the distance.

To better quantify how well we reconstruct the ratio, we have also considered $\Delta$ the mean
over the track of the relative error of $r_i$.

\begin{equation}
\Delta=\frac{1}{N}\sum_1^N\frac{|r_i-r_i^0|}{r_i^0}
\end{equation}

$\Delta$ values of each of the 100 noise realisation are shown as well as function of the distance
on Figure \ref{fig:s20results}. For a source located up to $\sim$\unit[9]{kpc} the relative error
remains smaller than 20\%.
At small distances, $\Delta$ is small but not null. This reflects the approximation of the model used for $r$.
It is nevertheless remarkable that, on average, one can reconstruct the ratio time series with a good
precision at distance up to $\sim$ 9 kpc for this particular waveform, with $coverage$ value
larger than 80\%. There are few noise realizations for a source located at $<$ \unit[9]{kpc} for which
$\Delta$ takes large values, indicating that the method start failing to reconstruct with accuracy the ratio.

\begin{figure}
  \centering
  \begin{tabular}{c}
    \includegraphics[width=0.5\textwidth]{plots/s20--SFHo_covpbb_boxplot_aLIGO} \\
    \includegraphics[width=0.5\textwidth]{plots/s20--SFHo_error_boxplot_aLIGO} \\
  \end{tabular}

 \caption{Boxplots of the $coverage$ (upper panel) and $\Delta$ (lower panel) for {\texttt s20S} signal embedded in aLIGO noise at different distances from the Earth. 100 noise realizations are considered for each distance.}
  \label{fig:s20results}
\end{figure}

We have tested that the method does not depend on features of {\texttt s20S} using the 7
other waveforms of the {\it test set} described in section \ref{sec:simulations} covering
a large range of progenitor masses.

Figure \ref{fig:aLIGO_cov_allwvf} shows that apart from {\texttt s11} and to a lesser extent
{\texttt s20S}, the ratio is well reconstructed for all waveforms up to $\sim$ 15kpc. In an
effort to better determine the maximal distance of the source at which we can reconstruct the ratio
we have run 100 simulations without injecting a signal and have measured $coverage$ for the
reconstructed ratios.
The median of $coverage$ as well as the 95 quartile are shown on Figure \ref{fig:aLIGO_cov_allwvf}.
The noise only median value is null in this case, but it can be different from zero because
the g-mode reconstruction algorithm is looking for a continuously frequency increasing track
in the spectrogram, starting between 0 and 200 Hz, where we expect the GW signal to be.
This is enhancing the probability of overlap. This effects explains why outliers can reach
values as high as 80\%.
Figure \ref{fig:aLIGO_prec_allwvf} shows $\Delta$ as function of the distance for the same signals
as well as the result when only noise is considered. 
In Table \ref{tab:results} we are reporting the distance $d_r$ at which $coverage$ median is lower
than 95\% of the noise only values. We have checked that $coverage$ and $\Delta$ provide similar
values.
These numbers are an estimate of the order of magnitude of the source maximal distance at which a
reconstruction of the ratio could be possible with current GW detectors.
They are also upper limits as we are taking into account the detector antenna response in our
simulation but consider the source is optimally oriented. 
Table \ref{tab:results} reports also $d_r$ for the Advanced Virgo detector at design sensitivity.
Results are very similar to aLIGO, despite the detector sensitivity differences.
Note that Table \ref{tab:results} provides the distance at which one could detect a source
optimally oriented with a matched filter signal-to-noise ratio of 13.

>>>>

To estimate how accurately we can infer the time evolution of $r(t)$ in the GW data of a single
detector we inject the GW signal for model {\texttt s20S} into 
100 Gaussian noise realisations whose power spectral density (PSD) follows the aLIGO
spectrum~\cite{aLIGOsens:2018}. The PSD of aLIGO and AdV at design sensitivities, along with those of planned third-generation detectors, are shown in Figure~\ref{fig:spectrum}.

\begin{figure}[t]
 \centering
 \includegraphics[width=0.5\textwidth]{plots/spectrum}
 \caption{Amplitude spectral density of the GW detectors aLIGO and AdV at design sensitivity and that of the proposed third-generation detectors Cosmic Explorer and Einstein Telescope. Einstein Telescope sensitivity curve ET-B is obtained pushing second-generation detector technology at its limit. ET-C and ET-D sensitivity curves correspond to a detector configuration where a low-power cryogenic low-frequency interferometer and a high-power room temperature high-frequency interferometer are sharing the same infrastructure~\cite{Hild_2011}. Cosmic Explorer design sensitivity will be achieved in two stages. Stage 1 (CE1) is expected to use the technology developed for the ``A+'' upgrade to aLIGO but scaled up to a 40 km detector while stage 2 (CE2) will implement state-of-the-art technology to decrease quantum and thermal noises~\cite{reitze2019cosmic}. } \label{fig:spectrum}
\end{figure}

We cover a large range of distances for which a CCSN detection in second-generation GW detectors is feasible. We assume that the source is optimally oriented with respect to our single detector. Moreover, we also assume that a CCSN GW signal has been identified in the data and that the beginning of the signal is known within {$\mathcal O$}(10 ms). The data (signal embedded in noise) are whitened using the function {\tt prewhiten} of the R-package {\tt TSA}. An auto-regressive model with a maximum of maximal 100 coefficients is used.

\begin{figure}[t]
  \centering
  \begin{tabular}{c}
    \includegraphics[width=0.5\textwidth]{plots/s20--SFHo_covpbb_boxplot_aLIGO} \\
    \includegraphics[width=0.5\textwidth]{plots/s20--SFHo_error_boxplot_aLIGO} \\
  \end{tabular}

 \caption{Boxplots of the $coverage$ (upper panel) and $\Delta$ (lower panel) for {\texttt s20S} signal embedded in aLIGO noise at different distances from the Earth. 100 noise realizations are considered for each distance. \tf{We should explain a bit more the symbols and colours that appear in the plots.}}
  \label{fig:s20results}
\end{figure}

For each of the noise realizations we reconstruct the ratio time series {$r_i$} of length $N$ starting from the left side of the spectrogram and constraining the beginning of the track to be smaller than \unit[200]{Hz}. The reconstructed ratio is then compared to the ``true'' ratio {$r_i^0$} derived from the PNS mass and radius computed from the {\texttt s20S} simulation. The top panel of Fig.~\ref{fig:s20results} shows the distribution of the fraction of the ratio values {$r_i^0$} that fall within the 95\% confidence interval of {$r_i$} as a function of the distance to the source. This quantity, hence, gives information about the {\it coverage} of the reconstructed ratio. The coverage takes maximum values when the source is located within a few kpc and then decreases with the distance.

To better quantify how well we reconstruct the ratio we also consider  the mean of the relative error of $r_i$ along 
the track of the spectrogram, $\Delta$,  
\begin{equation}
\Delta=\frac{1}{N}\sum_1^N\frac{|r_i-r_i^0|}{r_i^0}\,.
\end{equation}
The values of $\Delta$ for each of the 100 noise realizations are shown as a function of the distance
in the bottom panel of Fig.~\ref{fig:s20results}. For a source located up to $\sim$\unit[9]{kpc} the relative error
remains smaller than 20\%. At closer distances $\Delta$ is smaller but it does not vanish. This reflects the approximate nature of the model used for $r$. It is nevertheless remarkable that, on average, one can reconstruct the ratio time series with a good
precision up to distances of $\sim$ 9 kpc (for this particular waveform) with a coverage value
larger than 80\%. We note that there are a few noise realizations for distances below 9 kpc for which
$\Delta$ takes large values, indicating that the method fails to accurately reconstruct the ratio in those case. \tf{Do we know why that is the case? What is intrinsically different from one noise realisation to another that makes the method more prone to failure? I think we should provide some explanation here, if we have it.}

We have tested that the method does not depend on the specific features of the waveform of model {\texttt s20S} by repeating the procedure for the remaining seven waveforms of the {\it test set} described in Section \ref{sec:simulations} covering
a large range of progenitor masses. Figure \ref{fig:aLIGO_cov_allwvf} shows that apart from {\texttt s11} and to a lesser extent
{\texttt s20S}, the ratio is well reconstructed for all waveforms up to $\sim$ 15kpc. In an
effort to better determine the maximal distance of the source at which we can reconstruct the ratio
we have run 100 simulations without injecting a signal and have measured $coverage$ for the
reconstructed ratios.
The median of $coverage$ as well as the 95 quartile are shown on Figure \ref{fig:aLIGO_cov_allwvf}.
The noise only median value is null in this case, but it can be different from zero because
the g-mode reconstruction algorithm is looking for a continuously frequency increasing track
in the spectrogram, starting between 0 and 200 Hz, where we expect the GW signal to be.
This is enhancing the probability of overlap. This effects explains why outliers can reach
values as high as 80\%.
Figure \ref{fig:aLIGO_prec_allwvf} shows $\Delta$ as function of the distance for the same signals
as well as the result when only noise is considered. 
In Table \ref{tab:results} we are reporting the distance $d_r$ at which $coverage$ median is lower
than 95\% of the noise only values. We have checked that $coverage$ and $\Delta$ provide similar
values.
These numbers are an estimate of the order of magnitude of the source maximal distance at which a
reconstruction of the ratio could be possible with current GW detectors.
They are also upper limits as we are taking into account the detector antenna response in our
simulation but consider the source is optimally oriented. 
Table \ref{tab:results} reports also $d_r$ for the Advanced Virgo detector at design sensitivity.
Results are very similar to aLIGO, despite the detector sensitivity differences.
Note that Table \ref{tab:results} provides the distance at which one could detect a source
optimally oriented with a matched filter signal-to-noise ratio of 13.
